---
title: "Individual preprocessing"
---



```{r}
suppressPackageStartupMessages({
  library(Signac)
  library(Seurat)
  library(tidyverse)
  library(GenomicRanges)
  library(EnsDb.Mmusculus.v79)
  library(BSgenome.Mmusculus.UCSC.mm10)
  library(DoubletFinder)
  library(JASPAR2020)
  library(TFBSTools)
  library(harmony)
})

here::i_am("preprocess_individual.Rmd")
set.seed(1)
```


```{r}
# gene annotations
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79)
seqlevels(annotations) <- paste0('chr', seqlevels(annotations))
genome(annotations) <- "mm10"
```


```{r}
neo_data <- Read10X_h5("data/neo/filtered_feature_bc_matrix.h5")
tac_data <- Read10X_h5("data/4t/filtered_feature_bc_matrix.h5")
doa_data <- Read10X_h5("doa_data/filtered_feature_bc_matrix.h5")

neo_rna <- neo_data$`Gene Expression`
neo_atac <- neo_data$Peaks

tac_rna <- tac_data$`Gene Expression`
tac_atac <- tac_data$Peaks

doa_rna <- doa_data$`Gene Expression`
doa_atac <- doa_data$Peaks

# standard chromosomes only
keep_std_chr <- function(atac_counts) {
  grange.counts <- StringToGRanges(rownames(atac_counts), sep = c(":", "-"))
  grange.use <- seqnames(grange.counts) %in% standardChromosomes(grange.counts)
  atac_counts <- atac_counts[as.vector(grange.use), ]
  
  return(atac_counts)
}

neo_atac <- keep_std_chr(neo_atac)
tac_atac <- keep_std_chr(tac_atac)
doa_atac <- keep_std_chr(doa_atac)

# create seurat objects
neo_obj <- CreateSeuratObject(neo_rna, project = "neo")
tac_obj <- CreateSeuratObject(tac_rna, project = "tac")
doa_obj <- CreateSeuratObject(doa_rna, project = "doa")

neo_obj[["percent_mt"]] <- PercentageFeatureSet(neo_obj, pattern = "^mt-")
tac_obj[["percent_mt"]] <- PercentageFeatureSet(tac_obj, pattern = "^mt-")
doa_obj[["percent_mt"]] <- PercentageFeatureSet(doa_obj, pattern = "^mt-")

neo_obj[["condition"]] <- "Neonate"
tac_obj[["condition"]] <- "TAC"
doa_obj[["condition"]] <- ifelse(endsWith(colnames(doa_obj), c("1", "2")), "4M-H", "28M")
```


# Create common peak set
```{r}
neo_ranges <- StringToGRanges(rownames(neo_atac), sep = c(":", "-"))
tac_ranges <- StringToGRanges(rownames(tac_atac), sep = c(":", "-"))
doa_ranges <- StringToGRanges(rownames(doa_atac), sep = c(":", "-"))

combined_peaks <- IRanges::reduce(x = c(neo_ranges, tac_ranges, doa_ranges))

neo_fragment <- CreateFragmentObject(path = "data/neo/atac_fragments.tsv.gz", cells = colnames(neo_atac))
neo_atac_count <- FeatureMatrix(fragments = neo_fragment, features = combined_peaks, process_n = 4000) 
neo_obj[["ATAC"]] <- CreateChromatinAssay(neo_atac_count, fragments = neo_fragment, annotation = annotations)

tac_fragment <- CreateFragmentObject(path = "data/4t/atac_fragments.tsv.gz", cells = colnames(tac_atac))
tac_atac_count <- FeatureMatrix(fragments = tac_fragment, features = combined_peaks, process_n = 4000) 
tac_obj[["ATAC"]] <- CreateChromatinAssay(tac_atac_count, fragments = tac_fragment, annotation = annotations)

doa_fragment <- CreateFragmentObject(path = "doa_data/atac_fragments.tsv.gz", cells = colnames(doa_atac))
doa_atac_count <- FeatureMatrix(fragments = doa_fragment, features = combined_peaks, process_n = 4000) 
doa_obj[["ATAC"]] <- CreateChromatinAssay(doa_atac_count, fragments = doa_fragment, annotation = annotations)
```


```{r}
saveRDS(neo_obj, file = "/scratch/kwon.so/neo_obj.rds")
saveRDS(tac_obj, file = "/scratch/kwon.so/tac_obj.rds")
saveRDS(doa_obj, file = "/scratch/kwon.so/doa_obj.rds")

neo_obj <- readRDS("/scratch/kwon.so/neo_obj.rds")
tac_obj <- readRDS("/scratch/kwon.so/tac_obj.rds")
doa_obj <- readRDS("/scratch/kwon.so/doa_obj.rds")
```


# QC
```{r}
DefaultAssay(neo_obj) <- "ATAC"
neo_obj <- NucleosomeSignal(neo_obj)
neo_obj <- TSSEnrichment(neo_obj)

VlnPlot(object = neo_obj,
        features = c("nCount_RNA", "nCount_ATAC", "TSS.enrichment", "nucleosome_signal", "percent_mt"),
        ncol = 4,
        pt.size = 0,
        y.max = NULL)

# filtering low quality cells 
neo_obj <- subset(
    x = neo_obj,
    subset = nCount_RNA < 20000 & nCount_RNA > 200 &
    nCount_ATAC < 8000 & nCount_ATAC > 20 &
    nucleosome_signal < 2 &
    TSS.enrichment > 1 &
    percent_mt < 20
)
```

```{r}
DefaultAssay(tac_obj) <- "ATAC"
tac_obj <- NucleosomeSignal(tac_obj)
tac_obj <- TSSEnrichment(tac_obj)

VlnPlot(object = tac_obj,
        features = c("nCount_RNA", "nCount_ATAC", "TSS.enrichment", "nucleosome_signal", "percent_mt"),
        ncol = 4,
        pt.size = 0,
        y.max = NULL)

## filtering low quality cells
tac_obj <- subset(
  x = tac_obj,
  subset = nCount_RNA < 15000 & nCount_RNA > 200 &
  nCount_ATAC < 20000 & nCount_ATAC > 100 &
  nucleosome_signal < 2 &
  TSS.enrichment > 1 &
  percent_mt < 20
)
```

```{r}
DefaultAssay(doa_obj) <- "ATAC"
doa_obj <- NucleosomeSignal(doa_obj)
doa_obj <- TSSEnrichment(doa_obj)

VlnPlot(object = doa_obj,
        features = c("nCount_RNA", "nCount_ATAC", "TSS.enrichment", "nucleosome_signal", "percent_mt"),
        ncol = 4,
        pt.size = 0,
        y.max = NULL)

## filtering low quality cells
doa_obj <- subset(
  x = doa_obj,
  subset = nCount_RNA < 8000 & nCount_RNA > 200 &
  nCount_ATAC < 8500 & nCount_ATAC > 100 &
  nucleosome_signal < 2 &
  TSS.enrichment > 1 &
  percent_mt < 20
)
```


```{r}
saveRDS(neo_obj, file = "/scratch/kwon.so/filtered_neo_obj.rds")
saveRDS(tac_obj, file = "/scratch/kwon.so/filtered_tac_obj.rds")
saveRDS(doa_obj, file = "/scratch/kwon.so/filtered_doa_obj.rds")

neo_obj <- readRDS("/scratch/kwon.so/filtered_neo_obj.rds")
tac_obj <- readRDS("/scratch/kwon.so/filtered_tac_obj.rds")
doa_obj <- readRDS("/scratch/kwon.so/filtered_doa_obj.rds")
```


# Processing
```{r}
process_atac <- function(seu_obj, cut_off = 20) {
  DefaultAssay(seu_obj) <- "ATAC"
  
  seu_obj <- RunTFIDF(seu_obj)
  seu_obj <- FindTopFeatures(seu_obj, min.cutoff = cut_off)
  seu_obj <- RunSVD(seu_obj)
  
  return(seu_obj)
}

process_rna <- function(seurat_object, cluster_res = 0.5, dims = 1:30) {
  DefaultAssay(seurat_object) <- "RNA"
  
  seurat_object <- NormalizeData(seurat_object)
  seurat_object <- FindVariableFeatures(seurat_object)
  seurat_genes <- rownames(seurat_object)
  seurat_object <- ScaleData(seurat_object, features = seurat_genes)
  seurat_object <- RunPCA(seurat_object)
  seurat_object <- FindNeighbors(seurat_object, dims = dims, reduction = "pca")
  seurat_object <- FindClusters(seurat_object, resolution = cluster_res)
  seurat_object <- RunUMAP(seurat_object, dims = dims, reduction = "pca")
  
  return(seurat_object)
}

neo_obj <- process_atac(neo_obj)
neo_obj <- process_rna(neo_obj)

tac_obj <- process_atac(tac_obj)
tac_obj <- process_rna(tac_obj)

doa_obj <- process_atac(doa_obj)
doa_obj <- process_rna(doa_obj)
```


```{r}
saveRDS(neo_obj, file = "/scratch/kwon.so/processed_neo_obj.rds")
saveRDS(tac_obj, file = "/scratch/kwon.so/processed_tac_obj.rds")
saveRDS(doa_obj, file = "/scratch/kwon.so/processed_doa_obj.rds")

neo_obj <- readRDS("/scratch/kwon.so/processed_neo_obj.rds")
tac_obj <- readRDS("/scratch/kwon.so/processed_tac_obj.rds")
doa_obj <- readRDS("/scratch/kwon.so/processed_doa_obj.rds")
```


# Doublet detection and removal
```{r}
doublet_detect <- function(seu_obj, pcs = 1:10, expected_doublets = 0.05) {
  DefaultAssay(seu_obj) <- "RNA"
  
  sweep.res.list <- paramSweep(seu_obj, PCs = pcs, sct = FALSE)
  sweep.stats <- summarizeSweep(sweep.res.list, GT = FALSE)
  bcmvn <- find.pK(sweep.stats)
  
  pK_value <- as.numeric(as.character(bcmvn$pK[which.max(bcmvn$BCmetric)]))
  
  total_cells <- ncol(seu_obj)
  expected_doublet <- expected_doublets
  nExp_poi <- round(total_cells * expected_doublets)
  
  seu_obj <- doubletFinder(seu_obj, PCs = pcs, pN = 0.25, pK = pK_value, nExp = nExp_poi)
  
  seu_obj@meta.data <- seu_obj@meta.data %>%
    dplyr::rename("multiplet" = paste0("DF.classifications_0.25_", pK_value, "_", nExp_poi))
  
  return(seu_obj)
}

### change expected doublets as required 
neo_obj <- doublet_detect(neo_obj, expected_doublets = 0.05)
tac_obj <- doublet_detect(tac_obj, expected_doublets = 0.05)
doa_obj <- doublet_detect(doa_obj, expected_doublets = 0.05)

# view doublets
table(neo_obj@meta.data$multiplet)
table(tac_obj@meta.data$multiplet)
table(doa_obj@meta.data$multiplet)

# filter doublets 
neo_obj <- subset(neo_obj, subset = multiplet == "Singlet")
tac_obj <- subset(tac_obj, subset = multiplet == "Singlet")
doa_obj <- subset(doa_obj, subset = multiplet == "Singlet")
```


```{r}
saveRDS(neo_obj, file = "/scratch/kwon.so/doublet_neo_obj.rds")
saveRDS(tac_obj, file = "/scratch/kwon.so/doublet_tac_obj.rds")
saveRDS(doa_obj, file = "/scratch/kwon.so/doublet_doa_obj.rds")

neo_obj <- readRDS("/scratch/kwon.so/doublet_neo_obj.rds")
tac_obj <- readRDS("/scratch/kwon.so/doublet_tac_obj.rds")
doa_obj <- readRDS("/scratch/kwon.so/doublet_doa_obj.rds")
```


# Merge and integrate 
```{r}
neo_obj[["dataset"]] <- "N"
tac_obj[["dataset"]] <- "T"
doa_obj[["dataset"]] <- "D"

DefaultAssay(neo_obj) <- "ATAC"
DefaultAssay(tac_obj) <- "ATAC"
DefaultAssay(doa_obj) <- "ATAC"

merged_seu <- merge(x = neo_obj, y = list(tac_obj, doa_obj))

# process ATAC
merged_seu <- RunTFIDF(merged_seu)
merged_seu <- FindTopFeatures(merged_seu, min.cutoff = 20)
merged_seu <- RunSVD(merged_seu)

# integrating ATAC assay 
DefaultAssay(merged_seu) <- "ATAC"

integration.anchors <- FindIntegrationAnchors(  
  object.list = list(neo_obj, tac_obj, doa_obj),
  anchor.features = rownames(tac_obj),
  reduction = "rlsi",
  dims = 2:30
)

integrated <- IntegrateEmbeddings(  
  anchorset = integration.anchors,
  reductions = merged_seu[["lsi"]],
  new.reduction.name = "integrated_lsi",
  dims.to.integrate = 1:30, k.weight = 50
)

integrated <- RunUMAP(integrated, reduction = "integrated_lsi", dims = 2:30, reduction.name = "atac.UMAP")
DimPlot(integrated, group.by = "condition", reduction = "atac.UMAP")


# process RNA
DefaultAssay(integrated) <- "RNA"

integrated <- NormalizeData(integrated)
integrated <- FindVariableFeatures(integrated)

all.genes <- rownames(integrated)
integrated <- ScaleData(integrated, features = all.genes)
integrated <- RunPCA(integrated)

# integrating RNA assay 
integrated <- RunHarmony(integrated, "dataset", reduction = "pca", reduction.save = "harmony", assay.use = "RNA")
integrated <- RunUMAP(integrated, reduction = "harmony", reduction.name = "harmony.UMAP", dims = 1:30)
DimPlot(integrated, reduction = "harmony.UMAP", group.by = "condition")

integrated[["RNA"]] <- JoinLayers(integrated[["RNA"]])
```


```{r}
saveRDS(integrated, file = "/scratch/kwon.so/integrated_individual.rds")

integrated <- readRDS("/scratch/kwon.so/integrated_individual.rds")
```


# WNN 
```{r}
integrated <- FindMultiModalNeighbors(integrated, 
                                      reduction.list = list("pca", "integrated_lsi"), 
                                      dims.list = list(1:30, 2:30))

integrated <- RunUMAP(integrated, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
integrated <- FindClusters(integrated, graph.name = "wsnn", algorithm = 3, verbose = FALSE)

DimPlot(integrated, reduction = "wnn.umap", label = T) + ggtitle("WNN")
```


```{r}
saveRDS(integrated, file = "/scratch/kwon.so/integrated_wnn.rds")
integrated <- readRDS("/scratch/kwon.so/integrated_wnn.rds")
```


# Annotate cell types ---------- do this again 

CM = 0, 4, 9, 10, 14, 28, 29
EC = 1, 5, 6, 11, 15, 21^, 24^, 26^, 32
FB = 2, 8, 12, 17, 25^, 27^, 31^
Macro = 3, 18, 20, 22^, 23^^
B = 16
T = 19
Peri = 7, 13

SMC/Peri = 30

CM, EC, FB, Macro, CM, EC, EC, Peri, FB, CM, CM, EC, FB, Peri, CM, EC, B, FB, Macro, T, Macro, EC^, Macro^, Macro^^, EC^, FB^, EC^, FB^, CM, CM, SMC/Peri, FB^, EC 
```{r}
DefaultAssay(integrated) <- "RNA"

marker_genes <- list(
  CM = c("Tnnt2", "Actc1", "Myh6", "Myh7", "Myl2", "Myl7"),
  EC = c("Pecam1", "Cdh5", "Vegfa", "Kdr", "Nos3"),
  SMC = c("Acta2", "Tagln", "Myh11", "Smtn"),
  FB = c("Col1a1", "Col3a1", "Postn", "Dcn", "Tcf21"),
  Macro = c("Adgre1", "Cd68", "Itgam", "Csf1r", "Mrc1"),
  B_cells = c("Cd19", "Cd79a", "Ms4a1"),   
  T_cells = c("Cd3e", "Cd4", "Cd8a", "Il7r"),
  Meso = c("Msln", "Wt1"),
  Pericytes = c("Pdgfrb", "Cspg4", "Anpep"),
  Mono = c("Cd14", "Ms4a7"),
  NK = c("Nkg7"),
  DC = c("Cst3")
)

DotPlot(integrated, features = marker_genes)

types <- c("CM", "EC", "FB", "Macro", "CM", "EC", "EC", "Peri", "FB", "CM", "CM", "EC", "FB", "Peri", "CM", "EC", "B", "FB", "Macro", "T", "Macro", "EC", "Macro", "Macro", "EC", "FB", "EC", "FB", "CM", "CM", "SMC/Peri", "FB", "EC")

names(types) <- levels(integrated)

new_clusters <- data.frame(cell_type = types) %>% rownames_to_column("cluster")

# rownames as a column (merging removes rownames) 
integrated@meta.data <- integrated@meta.data %>% 
  rownames_to_column("cell_barcodes")

# merging  
integrated@meta.data <- left_join(x = integrated@meta.data, y = new_clusters, 
                                by = join_by(seurat_clusters == cluster),
                                relationship = "many-to-one")

# set rownames back
integrated@meta.data <- integrated@meta.data %>% 
  column_to_rownames("cell_barcodes")

Idents(integrated) <- integrated$cell_type

DimPlot(integrated, label = T, reduction = "wnn.umap")#, group.by = "condition")
```


```{r}
saveRDS(integrated, file = "/scratch/kwon.so/preprocessed_individual.rds")
integrated <- readRDS("/scratch/kwon.so/preprocessed_individual.rds")
```


## save as h5seurat
```{r}

```









# Clean cell types more ------------------ TODO: need to finish this and integrate and so on. 
Subset CM, FB, EC, and Macro; Process each subset; Cluster them and do DotPlot with the markers again. 
```{r}
# subset cell types
cm_seu <- subset(integrated, subset = cell_type == "CM")
fb_seu <- subset(integrated, subset = cell_type == "FB")
ec_seu <- subset(integrated, subset = cell_type == "EC")
macro_seu <- subset(integrated, subset = cell_type == "Macro")

# process each subset
process_subset <- function(seu_obj) {
  # processing RNA
  DefaultAssay(seu_obj) <- "RNA"
  seu_obj <- NormalizeData(seu_obj)
  seu_obj <- FindVariableFeatures(seu_obj)
  seu_obj <- ScaleData(seu_obj)
  seu_obj <- RunPCA(seu_obj)
  
  # process ATAC
  DefaultAssay(seu_obj) <- "ATAC"
  seu_obj <- RunTFIDF(seu_obj)
  seu_obj <- FindTopFeatures(seu_obj)
  seu_obj <- RunSVD(seu_obj)
  
  # wnn
  seu_obj <- FindMultiModalNeighbors(seu_obj, reduction.list = list("pca", "lsi"), dims.list = list(1:30, 2:30))
  seu_obj <- RunUMAP(seu_obj, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
  seu_obj <- FindClusters(seu_obj, graph.name = "wsnn", resolution = 0.8, algorithm = 3)
  
  return(seu_obj)
}

cm_seu <- process_subset(cm_seu)
fb_seu <- process_subset(fb_seu)
ec_seu <- process_subset(ec_seu)
macro_seu <- process_subset(macro_seu)

# check wnn umaps
DimPlot(cm_seu, reduction = "wnn.umap", label = T) + ggtitle("CM")
DimPlot(fb_seu, reduction = "wnn.umap", label = T) + ggtitle("FB")
DimPlot(ec_seu, reduction = "wnn.umap", label = T) + ggtitle("EC")
DimPlot(macro_seu, reduction = "wnn.umap", label = T) + ggtitle("Macro")

# DimPlot(cm_seu, reduction = "wnn.umap", label = T, group.by = "condition") + ggtitle("CM")
# DimPlot(fb_seu, reduction = "wnn.umap", label = T, group.by = "condition") + ggtitle("FB")
# DimPlot(ec_seu, reduction = "wnn.umap", label = T, group.by = "condition") + ggtitle("EC")
# DimPlot(macro_seu, reduction = "wnn.umap", label = T, group.by = "condition") + ggtitle("Macro")

# dotplot of markers
DotPlot(cm_seu, features = marker_genes) # neonates = 7
DotPlot(fb_seu, features = marker_genes) # neonates = 8
DotPlot(ec_seu, features = marker_genes) # neonates = 14
DotPlot(macro_seu, features = marker_genes) # neonates = 8

# remove clusters that show multiple markers 
cm_seu <- subset(cm_seu, subset = !seurat_clusters %in% c(8, 9, 10, 11))
fb_seu <- subset(fb_seu, subset = !seurat_clusters %in% c(4, 7, 9, 10, 11, 13))
ec_seu <- subset(ec_seu, subset = !seurat_clusters %in% c(10, 11, 13, 15))
macro_seu <- subset(macro_seu, subset = !seurat_clusters %in% c(3, 5, 10, 11))

# merge cell types into one object
t_seu <- subset(integrated, subset = cell_type == "T")
b_seu <- subset(integrated, subset = cell_type == "B")
peri_seu <- subset(integrated, subset = cell_type == "Peri")
smc_seu <- subset(integrated, subset = cell_type == "SMC/Peri")

clean_seu <- merge(x = cm_seu, y = list(fb_seu, ec_seu, macro_seu, t_seu, b_seu, peri_seu, smc_seu))

# process clean_seu and integrate with harmony
## process RNA first 

## process ATAC next

## integrate both assay

## join layers

## perform wnn



##### testing ######
process_subset(clean_seu) #### do not use the function. integrate first and then do wnn

DefaultAssay(clean_seu) <- "RNA"

DimPlot(clean_seu)
```






